A performance comparison of Dask and Spark for data-intensive neuroimaging pipelines

### 2019-07-25 latex/paper.tex
fixed spark section

\subsubsection{Apache Spark} Apache Spark is a general-purpose Big Data engine.
Spark's core is written in Scala, but it also has a Java, an R, and a Python API
called PySpark. We used PySpark, given the popularity of Python in neuroscience,
although the serialization cost of Python objects to Scala might be a substantial
source of overhead. Spark provides in-memory computing, data locality, and lazy
evaluation. PySpark's fundamental data structure, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant and a parallel collection of data elements.
Dataframes, PySpark's API for tabular data, is an overlay of RDDs with similar
performance features. We used Spark's default scheduler, Spark standalone, although
YARN~\cite{vavilapalli2013apache} and Mesos~\cite{hindman2011mesos} are also
available. In the Spark standalone scheduler, a process called the \emph{master}
coordinates resources provisioned by \emph{worker} processes on the cluster. The
application is submitted to the \emph{driver} that in turn requests workers to the
master and dispatches tasks to them. A job is divided into stages to be excuted in a
different process onto the workers. Each stage's operation is represented as a
high-level task in the computation graph. The Spark standalone scheduler uses a FIFO
(First-In-First-Out) job scheduling policy. Processing jobs in order helps in
increasing the page hit rate. As our applications only have one job, the LIFO
(Last-In-First-Out) policy used for the stages is more pertinent \MD{Find better
phrasing} \TG{refer to the paper (\MD{Not sure which one})} \VHS{I wouldn't
explicitly state LIFO here. It's a bit complicated to be frank. Basically both Spark
and Dask opt to complete a chain of dependencies rather than start a bunch of ``new
chains''. And in this context, LIFO is great for the page hit rate, so you might want
to reword the prior sentence as well}. The Spark standalone scheduler has two
execution modes: client mode, where the driver runs in a dedicated process on the
cluster, and cluster mode, where the driver runs in a worker. We used client mode as
cluster mode is not available in PySpark. We used Apache Spark v2.4.0.

\subsubsection{Dask} Dask is a Big Data engine that is becoming increasingly popular
in the scientific Python ecosystem. Dask has five main APIs and data structures:
Array, Bag, DataFrame, Delayed, and Futures. All APIs were used in our experiments,
with the exception of Dask Dataframes, which is intended for use with tabular data.
Like Spark, all Dask APIs provide
in-memory computing, data locality, lazy evaluation, and fault tolerance by mean of
lineage. Dask's core is purely in Python and bypasses most of the
\href{https://docs.python.org/3/glossary.html#term-gil}{GIL} to provide parallelism.
We used Dask due to the common use of the scientific Python ecosystem in
neuroscience\VHS{If you add Dask Arrays to your experiments here, i'd mention that their
APIs might be particularly well suited for the processing of high resolution neuroimaging data}. Dask represents application tasks as Dask graphs, produced by a
user-facing API and executed by a scheduler. In Dask, operations generate multiple
tiny tasks in the computation graph allowing an easier representation of complex
algorithms. We used the \href{https://distributed.dask.org/en/latest/index.html}{Dask
Distributed} scheduler, although \href{https://github.com/dask/dask-yarn}{Dask-Yarn}
and \href{https://github.com/mrocklin/dask-mesos}{Dask-Mesos} is also available. In
the Dask Distributed scheduler, a process called \textit{dask-scheduler}
administrates the resources provided by \textit{workers} in the cluster. The
scheduler receives jobs from clients and assigns tasks to available workers. Like
Spark scheduler, Dask Distributed also use a LIFO policy, it is meant to reduce the
memory footprint of the application by finishing the computation of a branch in the
Dask graph before starting new ones. We used Dask v1.1.4.

@@

%%% INCREMENTATION %%%
%% worker
\subsection{Incrementation: Number of workers}
Figure~\ref{fig:inc_ms_worker} shows the makespan of the Incrementation application
for different numbers of workers and for different engines. The bars show the average
makespan over 3 repetitions while the error bars are the standard deviation. Overall,
there is no substantial makespan difference between the engines. Dask seems to have a
slight advantage over Spark, \SI{83.61}{\second} on average,
with Delayed and Futures being slightly better than Bags.

@@

%% Iterations
\subsection{Incrementation: Number of iterations}
Figure~\ref{fig:inc_ms_itr} shows the makespan of the application while
varying the number of iterations. Overall,  Spark and Dask APIs are once
again equivalent, although Dask Delayed and Futures are slightly faster
than Bags and Spark for 1 and 10 few iterations, and Futures are faster than
Delayed, Bags and Spark for 100 iterations. Differences remain minor
though.

In Figure~\ref{fig:inc_tt_itr}, the total execution time breakdown is
shown. We observe good scalability of all the engines with the number of
iterations.
% TODO

@@

\begin{figure*}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/spark_inc_gantt.png}
        \caption{Spark}\label{fig:inc_spark_gantt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/bag_inc_gantt.png}%
        \caption{Dask Bag}\label{fig:inc_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/delayed_inc_gantt.png}%
        \caption{Dask Delayed}\label{fig:inc_dask_delayed_gantt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/futures_inc_gantt.png}%
        \caption{Dask Futures}\label{fig:inc_dask_futures_gantt}
    \end{subfigure}
    \caption{Incrementation: Gantt chart with 125 blocks, 10 iterations, \SI{4}{\second}
delay, 8 workers, 8 threads/worker}\label{fig:inc_gantt}
\end{figure*}

@@

From Figure~\ref{fig:histo_tt_worker} the total time spend in each function is shown.
The computation time is significantly larger in Dask than Spark. This is potentially
due to Python's GIL preventing Dask to parallelize the computation on multiple threads.
Overall, the IO and overhead is comparable for all engines.

On Figure~\ref{fig:histo_tt_worker}, Dask engines have similar total execution time
when their number of workers varies, except for 8 worker where it incresase slightly.
On the other hand, Spark total execution time keeps increasing; especially at 8
workers. This is because Dask engines benefit more from additional workers.

@@

\subsection{Histogram: Number of blocks}
Figure~\ref{fig:histo_ms_block} shows the makespan for different block size.
Again, Spark is significantly faster then Dask engines and Bag is slightly faster
than Delayed. Overall, the engines do not react to change in block size.

In figure~\ref{fig:histo_tt_block} 
%TODO

@@

\subsection{Histogram: Gantt chart}

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/spark_histo_gantt.png}
        \caption{Spark}\label{fig:histo_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/bag_histo_gantt.png}%
        \caption{Dask Bag}\label{fig:histo_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/delayed_histo_gantt.png}%
        \caption{Dask Delayed}\label{fig:histo_dask_delayed_gantt}
    \end{subfigure}
    \caption{Histogram: Gantt chart with 125 blocks, 10 iterations, \SI{4}{\second}
delay, 8 workers, 8 threads/worker}\label{fig:histo_gantt}
\end{figure}

%%% HISTOGRAM (NumPy) %%%
\subsection{Histogram (NumPy): Number of workers}
Figure~\ref{fig:histo_np_ms_worker} shows the makespan for various amount fo workers.
Overall, the makespan for all engines does not change considerably with an increase 
in worker. This is beacuse the engines do not scale well.

On figure~\ref{fig:histo_np_ms_worker}, there is no substantial difference between
 all engines, althought, Delayed is slightly faster.

From Figure~\ref{fig:histo_np_tt_worker}, the total execution time spent in each
function is shown. The overhead and read time increase proportionally to the number
of workers thus increasing the total time. This is because (1) most worker read at
the same time due to a low computation-IO ratio and (2) \MD{Determine why the
overhead is so large compared to the pure Python implementation}

\begin{figure}[!t]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/histo_np_worker.png}%
        \caption{Histogram (NumPy) makespan}\label{fig:histo_np_ms_worker}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/histo_idle_np_worker.png}%
        \caption{Histogram (NumPy) total time}\label{fig:histo_np_tt_worker}
    \end{subfigure}
    \caption{125 blocks, 8 threads/worker}
\end{figure}

\subsection{Histogram (NumPy): Number of blocks}
Figure~\ref{fig:histo_np_ms_block} shows the makespan of the application for
different block size. Note that the workers can process up to 64 tasks in paralle thus
they are not used at full capacity with 30 blocks. Overall, there is no substantial
difference for all engines and block size.

In figure~\ref{fig:histo_np_tt_block}, shows the total time spent in each function.
Again, the overhead for 30 blocks is erroneous due a lower amount of task than thread
which result in higher overhead for thread based engines. Considering the overhead
error and that at 30 blocks the worker only use half their core, the  total time is
substantially the same for all block size. This is due lower block sizes having a
lower IO time, however, it is balanced by the resulting increase in overhead.

\begin{figure}[!t]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/histo_np_block.png}%
        \caption{Histogram (NumPy) makespan}\label{fig:histo_np_ms_block}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/histo_idle_np_block.png}%
        \caption{Histogram (NumPy) total time}\label{fig:histo_np_tt_block}
    \end{subfigure}
    \caption{8 workers, 8 threads/worker}
\end{figure}

\subsection{Histogram (NumPy): Gantt chart}

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/spark_histo_np_gantt.png}
        \caption{Spark}\label{fig:histo_np_spark_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/bag_histo_np_gantt.png}%
        \caption{Dask Bag}\label{fig:histo_np_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/delayed_histo_np_gantt.png}%
        \caption{Dask Delayed}\label{fig:histo_np_dask_delayed_gantt}
    \end{subfigure}
    \caption{Histogram (NumPy): Gantt chart with 125 blocks, 10 iterations, \SI{4}{\second}
delay, 8 workers, 8 threads/worker}\label{fig:histo_np_gantt}
\end{figure}

%%% BIDS EXAMPLE %%%
\subsection{BidsApp example: Number of workers}
Figure~\ref{fig:bids_ms_worker} shows the makespan of the application while varying
the amount of worker. Overall, there is no substantial difference between the
engines. The makespan scales well with the increase in workers however the scaling
reduces considerably when reaching 8 workers. This is due to the IO requirement
increasing with the number of workers --- more tasks are read concurrently and
the application becomes more data-intensive.

In Figure~\ref{fig:bids_tt_worker}, the total execution time of each function is
shown. Overall, there is no significant difference between the engines. Each engines
have similar total time independently of their number of worker. This is because the
engines scale almost linearly. The overhead increase slowly as the number of workers
increases. This is due to more communication between the scheduler and workers as
well as inter-worker communication.

@@

%% Baseline gantt chart
\subsection{BidsApp example: Gantt chart}
Figure~\ref{fig:bids_gantt} shows the Gantt charts obtained for each engine and API.
The Gantt charts are structured in two main parts: participant analysis (orange) and
group analysis (blue). The participant analysis tasks differ greatly in lenght. This
is due to inqual amount of sessions per subject to process. The group analysis is
similar for all engines and APIs. Overall, most of the overhead encountered results
from the transition between the two analysis. This is because the group analysis is
requires the results of the participant analysis to start.

\begin{figure*}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/spark_bids_gantt.png}
        \caption{Spark}\label{fig:bids_spark_gantt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/bag_bids_gantt.png}%
        \caption{Dask Bag}\label{fig:bids_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/delayed_bids_gantt.png}%
        \caption{Dask Delayed}\label{fig:bids_dask_delayed_gantt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/futures_bids_gantt.png}%
        \caption{Dask Futures}\label{fig:bids_dask_futures_gantt}
    \end{subfigure}
    \caption{BIDS App example: Gantt chart with 8 workers}\label{fig:bids_gantt}
\end{figure*}

>>>>

\subsubsection{Apache Spark} Apache Spark is a general-purpose Big Data engine.
While its core is written in Scala, Spark provides additional APIs in Java, R
and Python (PySpark). Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. This
data structure is the basis of Spark's other data structures, namely, DataFrames
and Datasets. Datasets are similar to RDDs but additionally use Spark SQL's
optimized execution to further improve performance. DataFrames, used to process
tabular data, Datasets where the data is organized into named-columns. While 
the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

As data transfers in BigData workflows are an important source of performance
bottlenecks, Spark incorporated
data locality, and additionally, introduced in-memory computing to the Big Data 
frameworks. Data locality, introduced by MapReduce~\cite{mapreduce}, schedules
tasks as close as possible to where the data is storage. In-memory computing ensures
data is maintained in memory whenever possible, as writting large amounts data to
slower storage devices may be costly. To reduce any unecessary communication
and computation, Spark also included lazy evaluation, which builds the entire
task graph prior to execution to determine what needs to be computed.

Spark is compatible with three different schedulers: Spark
standalone, YARN~\cite{vavilapalli2013apache} and Mesos~\cite{hindman2011mesos}. 
The Spark standalone is a simple default scheduler built into Spark. YARN is a
more complex scheduler primarily designed to schedule Hadoop-based workflows. 
Mesos, on the other hand, is not limited to Hadoop and can be used to schedule
a variety of different workflows. As scientific researchers would likely be
executing their workflows in HPC environments with neither YARN or Mesos
installed, we limit our focus to Spark's Standalone scheduler.

The Spark Standalone scheduler is composed of three main processes: the
\emph{master}, the \emph{workers} (\emph{slaves}, in Spark) and the
\emph{driver}. The \emph{master} coordinates resources provisioned by
\emph{worker} processes on the cluster. The application is submitted to the
\emph{driver} that in turn requests workers to the master and dispatches tasks
to them. A job is divided into stages to be excuted in a different process onto
the workers. Each stage's operation is represented as a high-level task in the computation
graph. The Spark standalone scheduler uses a FIFO (First-In-First-Out) job
scheduling policy. The Spark standalone scheduler has two
execution modes: client mode, where the driver runs in a dedicated process outside
of the Spark cluster, and cluster mode, where the driver runs in a worker. We used
client mode as cluster mode is not available in PySpark. We used Apache Spark v2.4.0.

Python is commonly selected as the programming language of choice in scientific
communities, especially for our use case of neuroscience, where numerous specialized
Python libraries exist to study the data. While serialization of 
Python to Java may lead to significant overheads, we chose to focus on Spark 
Python API due to its suitability for neuroimaging research.

\subsubsection{Dask} Dask is a Big Data engine that is becoming increasingly popular
in the scientific Python ecosystem. Dask has five main data structure APIs:
Array, Bag, DataFrame, Delayed, and Futures. All APIs were used in our experiments,
with the exception of Dask Dataframes, which is intended for use with tabular data.
Like Spark, all Dask APIs provide
in-memory computing, data locality, lazy evaluation, and fault tolerance by mean of
lineage. Dask's core is purely in Python and bypasses most of the
\href{https://docs.python.org/3/glossary.html#term-gil}{GIL} to provide parallelism.
We used Dask due to the common use of the scientific Python ecosystem in
neuroscience\VHS{If you add Dask Arrays to your experiments here, i'd mention that their
APIs might be particularly well suited for the processing of high resolution neuroimaging data}. Dask represents application tasks as Dask graphs, produced by a
user-facing API and executed by a scheduler. In Dask, operations generate multiple
tiny tasks in the computation graph allowing an easier representation of complex
algorithms. We used the \href{https://distributed.dask.org/en/latest/index.html}{Dask
Distributed} scheduler, although \href{https://github.com/dask/dask-yarn}{Dask-Yarn}
and \href{https://github.com/mrocklin/dask-mesos}{Dask-Mesos} is also available. In
the Dask Distributed scheduler, a process called \textit{dask-scheduler}
administrates the resources provided by \textit{workers} in the cluster. The
scheduler receives jobs from clients and assigns tasks to available workers. Like
Spark scheduler, Dask Distributed also use a LIFO policy, it is meant to reduce the
memory footprint of the application by finishing the computation of a branch in the
Dask graph before starting new ones. We used Dask v1.1.4.

@@

%%% INCREMENTATION %%%
%% worker
\subsection{Incrementation: Number of workers}
Figure~\ref{fig:inc_ms_worker} shows the makespan of the Incrementation application
for different numbers of workers and for different engines. The bars show the average
makespan over 5 repetitions while the error bars are the standard deviation. Overall,
there is no substantial makespan difference between the engines. Dask seems to have a
slight advantage over Spark, \SI{83.61}{\second} on average,
with Delayed and Futures being slightly better than Bags.

@@

%% Iterations
\subsection{Incrementation: Number of iterations}
Figure~\ref{fig:inc_ms_itr} shows the makespan of the application while
varying the number of iterations. Overall,  Spark and Dask APIs are once
again equivalent, although Dask Delayed and Futures are slightly faster
than Bags and Spark for 1 and 10 few iterations, and Fuures are faster than
Delayed, Bags and Spark for 100 iterations. Differences remain minor
though.

In Figure~\ref{fig:inc_tt_itr}, the total execution time breakdown is
shown. We observe good scalability of all the engines with the number of
iterations.

@@

\begin{figure*}[!htb]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/spark_inc_gantt.png}
        \caption{Spark}\label{fig:inc_spark_gantt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/bag_inc_gantt.png}%
        \caption{Dask Bag}\label{fig:inc_dask_bag_gantt}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/delayed_inc_gantt.png}%
        \caption{Dask Delayed}\label{fig:inc_dask_delayed_gantt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\columnwidth}
        \includegraphics[clip,width=\columnwidth]{images/futures_inc_gantt.png}%
        \caption{Dask Futures}\label{fig:inc_dask_futures_gantt}
    \end{subfigure}
    \caption{Gantt chart with 125 blocks, 10 iterations, \SI{4}{\second} delay, 8
    workers, 8 threads/worker}\label{fig:inc_gantt}
\end{figure*}

@@

From Figure~\ref{fig:histo_tt_worker} the total time spend in each function is shown.
The computation time is significantly larger in Dask than Spark. This is potentially
due to Python's GIL preventing Dask to parallelize the computation on multiple thread.
Overall, the IO and overhead is comparable for all engines.

@@

\subsection{Histogram: Number of blocks}
Figure~\ref{fig:histo_ms_block}

@@

%%% BIDS EXAMPLE %%%
\subsection{BidsApp example: Number of workers}
Figure~\ref{fig:bids_ms_worker} shows the makespan of the application while varying
the amount of worker. Overall, there is no substantial difference between the
engines however Futures are always slightly faster. The performance difference between
Futures and the other APIs increases with the number of workers. The makespan scales
well with the increase in workers however the scaling reduces considerably when
reaching 8 workers. This is due to the IO requirement increasing with the number of
workers, since more tasks are read concurrently, and the application becomes more
data-intensive.

In Figure~\ref{fig:bids_tt_worker}, the total execution time of each function is
shown. \MD{Maybe mention why Spark overhead is so big.} Overall, the computation time
for the group analysis is equal for all engines and numbers of workers. Bag and
Delayed takes much longer to compute the participant analysis with 8 workers,
otherwise, the makespan is similar for all engines and numbers of workers. The overhead
increases as the number of workers increases. \MD{Check regression}

\MD{Depending on how the overhead is calculated spark overhead is either minimal or
gigantic. It is hard to determine which method should be used to calculate the
overhead for this application. For this application, since Spark has 2 job this
causes Spark to have different worker process for the different job and thus some job
have overlaping process ID while other are uniques (See
Figure~\ref{fig:bids_spark_gantt}).}

@@

%% Baseline gantt chart
\subsection{BidsApp example: Gantt chart}
Figure~\ref{fig:bids_gantt} shows the Gantt charts obtained for each engine and API.
The Gantt charts are structured in three main parts: participant analysis (orange),
group analysis (blue), and the transition in between.\MD{rephrase} The Gantt chart
for Spark seems to have more workers however it does not. This is due to the
computation of the participant and group analysis on two different jobs with their
worker processes; usually with different process id. The tasks differ considerably in
length. This is due to the high IO requirement of the application. During the
participant analysis phase, Spark and Futures have less overhead than Delayed and
Bag. At the end of the participant analysis phase, Futures and Delayed has less
overhead than Bag while Spark has way more than Dask engine. The group analysis is
similar for all engines and APIs.

% \begin{figure*}[!htb]
%     \centering
%     \begin{subfigure}[b]{\columnwidth}
%         \includegraphics[clip,width=\columnwidth]{images/spark_bids_gantt.png}
%         \caption{Spark}\label{fig:bids_spark_gantt}
%     \end{subfigure}
%     \hfill
%     \begin{subfigure}[b]{\columnwidth}
%         \includegraphics[clip,width=\columnwidth]{images/bag_bids_gantt.png}%
%         \caption{Dask Bag}\label{fig:bids_dask_bag_gantt}
%     \end{subfigure}
%     \\
%     \begin{subfigure}[b]{\columnwidth}
%         \includegraphics[clip,width=\columnwidth]{images/delayed_bids_gantt.png}%
%         \caption{Dask Delayed}\label{fig:bids_dask_delayed_gantt}
%     \end{subfigure}
%     \hfill
%     \begin{subfigure}[b]{\columnwidth}
%         \includegraphics[clip,width=\columnwidth]{images/futures_bids_gantt.png}%
%         \caption{Dask Futures}\label{fig:bids_dask_futures_gantt}
%     \end{subfigure}
%     \caption{Gantt chart with 8 workers}\label{fig:bids_gantt}
% \end{figure*}
